---
title: "CSE514PA1Code"
author: "Wenjia Zhang, 510260"
date: "2023-10-08"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#read the data
library(readxl)
Concrete_Data <- read_excel("/Users/Mika/Desktop/CSE514/PA1/Concrete_Data.xls")
```

## Data pre-processing

```{r}
#replace variable name in easier way
raw_data <- Concrete_Data
names(raw_data) <- c("Cement",
                 "Blast_Furnace_Slag",
                 "Fly_Ash",
                 "Water",
                 "Superplasticizer ",
                 "Coarse_Aggregate",
                 "Fine_Aggregate",
                 "Age",
                 "Concrete_strength")
```

```{r}
#visualize the raw data
boxplot(raw_data, 
        names=c('x1','x2','x3','x4','x5','x6','x7','x8','y'), 
        main = " The Distribution of Feature Values before Processing"  )
```

```{r}
#Standardize the data
new_data <- raw_data
new_data <- scale(new_data)

#save as data frame `data1`
new_data <- as.data.frame(new_data)

#visualize the data after standardization
boxplot(new_data, 
        names=c('x1','x2','x3','x4','x5','x6','x7','x8','y'),
        main = "The Distribution of Feature Values after Standardization")
```


## Linear Regression
```{r}
#Randomly select 900 training data and 130 testing data
set.seed(12345)
#raw data
raw_train_index <- sample(1:nrow(raw_data), size = 900)
raw_test_index <- setdiff(1:nrow(raw_data), raw_train_index)
raw_train <- raw_data[raw_train_index, ]
raw_test <- raw_data[raw_test_index, ]

#pre-processed data
new_train_index <- sample(1:nrow(new_data), size = 900)
new_test_index <- setdiff(1:nrow(new_data), new_train_index)
new_train <- new_data[new_train_index, ]
new_test <- new_data[new_test_index, ]
```



### Univariate linear regression

```{r}
lm <- function(x_train, y_train, x_test, y_test, iter = 1000, alpha){
  # Initialize parameters
  m <- 0
  b <- 0
  
  n <- length(x_train)
  
  # calculate gradient descent
  for(i in 1:iter){
    y_pred <- m * x_train + b
    grad_m <- (-2/n) * sum(x_train * (y_train - y_pred))  
    grad_b <- (-2/n) * sum(y_train - y_pred)  
    m <- m - alpha * grad_m  # Update m
    b <- b - alpha * grad_b  # Update b
  }
  
  # predictions
  y_pred_train <- m * x_train + b
  y_pred_test <- m * x_test + b
  
  # Calculate MSE, Variance and R-squared for training data
  MSE_train <- sum((y_train - y_pred_train)^2) / n
  Var_train <- sum((y_train - mean(y_train))^2) / n
  r_squared_train <- 1 - (MSE_train / Var_train)
  
  # Calculate MSE, Variance and R-squared for test data
  n_test = length(x_test)
  MSE_test <- sum((y_test - y_pred_test)^2) / n_test
  Var_test <- sum((y_test - mean(y_test))^2) / n_test
  r_squared_test <- 1 - (MSE_test / Var_test)

  # Return the coefficients, variance explained and predictions(for plot)
  list(coefficients = c(m = m, b = b), VE_train = r_squared_train, VE_test = r_squared_test, y_pred_train= y_pred_train)
}
```


#### Raw Data

##### 1. Cement
```{r}
lm1_raw <- lm(raw_train$Cement,raw_train$Concrete_strength, raw_test$Cement, raw_test$Concrete_strength, iter = 1000, alpha = 0.00001 )

#coefficients and the VE
lm1_raw[1:3]

#plot the data
plot(raw_train$Cement, raw_train$Concrete_strength, xlab = 'Cement',
       ylab ="strength")
lines(raw_train$Cement, lm1_raw$y_pred_train, col = "red")
```





##### 2. Blast_Furnace_Slag
```{r}
#raw data
lm2_raw = lm(raw_train$Blast_Furnace_Slag,raw_train$Concrete_strength, raw_test$Blast_Furnace_Slag, raw_test$Concrete_strength, iter = 1000, a = 0.00001 )

#coefficients and the VE
lm2_raw[1:3]
```

```{r}
#plot the data
plot(raw_train$Blast_Furnace_Slag, raw_train$Concrete_strength, xlab = 'Blast Furnace Slag',
       ylab ="strength")
lines(raw_train$Blast_Furnace_Slag, lm2_raw$y_pred_train, col = "red")
```

##### 3. Fly Ash
```{r}
#raw data
lm3_raw = lm(raw_train$Fly_Ash,raw_train$Concrete_strength, raw_test$Fly_Ash, raw_test$Concrete_strength, iter = 1000, a = 0.00001 )

lm3_raw[1:3]
```
```{r}
#plot the data
plot(raw_train$Fly_Ash, raw_train$Concrete_strength, xlab = 'Fly Ash',
       ylab ="strength")
lines(raw_train$Fly_Ash, lm3_raw$y_pred_train, col = "red")
```

##### 4. Water
```{r}
#raw data
lm4_raw = lm(raw_train$Water,raw_train$Concrete_strength, raw_test$Water, raw_test$Concrete_strength, iter = 1000, a = 0.00001 )

lm4_raw[1:3]
```
```{r}
#plot the data
plot(raw_train$Water, raw_train$Concrete_strength, xlab = 'Water',
       ylab ="strength")
lines(raw_train$Water, lm4_raw$y_pred_train, col = "red")
```

##### 5. Superplasticizer
```{r}
lm5_raw = lm(raw_train$`Superplasticizer `,raw_train$Concrete_strength, raw_test$`Superplasticizer `, raw_test$Concrete_strength, iter = 1000, a = 0.01 )

lm5_raw[1:3]
```
```{r}
#plot the data
plot(raw_train$`Superplasticizer `, raw_train$Concrete_strength, xlab = 'Superplasticizer',
       ylab ="strength")
lines(raw_train$`Superplasticizer `, lm5_raw$y_pred_train, col = "red")
```

##### 6. Coarse Aggregate
```{r}
lm6_raw = lm(raw_train$Coarse_Aggregate,raw_train$Concrete_strength, raw_test$Coarse_Aggregate, raw_test$Concrete_strength, iter = 1000, a = 0.0000001 )

lm6_raw[1:3]
```
```{r}
#plot the data
plot(raw_train$Coarse_Aggregate, raw_train$Concrete_strength, xlab = 'Coarse Aggregate',
       ylab ="Strength")
lines(raw_train$Coarse_Aggregate, lm6_raw$y_pred_train, col = "red")
```

##### 7. Fine Aggregate
```{r}
lm7_raw = lm(raw_train$Fine_Aggregate,raw_train$Concrete_strength, raw_test$Fine_Aggregate, raw_test$Concrete_strength, iter = 1000, a = 0.000001 )

lm7_raw[1:3]
```

```{r}
#plot the data
plot(raw_train$Fine_Aggregate, raw_train$Concrete_strength, xlab = 'Fine Aggregate',
       ylab ="Strength")
lines(raw_train$Fine_Aggregate, lm7_raw$y_pred_train, col = "red")
```

##### 8. Age
```{r}
lm8_raw = lm(raw_train$Age,raw_train$Concrete_strength, raw_test$Age, raw_test$Concrete_strength, iter = 1000, a = 0.0001 )

lm8_raw[1:3]
```
```{r}
#plot the data
plot(raw_train$Age, raw_train$Concrete_strength, xlab = 'Age',
       ylab ="Strength")
lines(raw_train$Age, lm8_raw$y_pred_train, col = "red")
```

#### Pre-processed Data


##### 1. Cement
```{r}
lm1_new <- lm(new_train$Cement,new_train$Concrete_strength, new_test$Cement, new_test$Concrete_strength, iter = 1000, alpha = 0.01 )

#coefficients and the VE
lm1_new[1:3]

#plot the data
plot(new_train$Cement, new_train$Concrete_strength, xlab = 'Cement',
       ylab ="strength")
lines(new_train$Cement, lm1_new$y_pred_train, col = "red")
```


##### 2. Blast_Furnace_Slag
```{r}
#new data
lm2_new = lm(new_train$Blast_Furnace_Slag,new_train$Concrete_strength, new_test$Blast_Furnace_Slag, new_test$Concrete_strength, iter = 1000, a = 0.001 )

#coefficients and the VE
lm2_new[1:3]
```

```{r}
#plot the data
plot(new_train$Blast_Furnace_Slag, new_train$Concrete_strength, xlab = 'Blast Furnace Slag',
       ylab ="strength")
lines(new_train$Blast_Furnace_Slag, lm2_new$y_pred_train, col = "red")
```

##### 3. Fly Ash
```{r}
#new data
lm3_new = lm(new_train$Fly_Ash,new_train$Concrete_strength, new_test$Fly_Ash, new_test$Concrete_strength, iter = 1000, a = 0.001 )

lm3_new[1:3]
```
```{r}
#plot the data
plot(new_train$Fly_Ash, new_train$Concrete_strength, xlab = 'Fly Ash',
       ylab ="strength")
lines(new_train$Fly_Ash, lm3_new$y_pred_train, col = "red")
```

##### 4. Water
```{r}
#new data
lm4_new = lm(new_train$Water,new_train$Concrete_strength, new_test$Water, new_test$Concrete_strength, iter = 1000, a = 0.01 )

lm4_new[1:3]
```
```{r}
#plot the data
plot(new_train$Water, new_train$Concrete_strength, xlab = 'Water',
       ylab ="strength")
lines(new_train$Water, lm4_new$y_pred_train, col = "red")
```

##### 5. Superplasticizer
```{r}
lm5_new = lm(new_train$`Superplasticizer `,new_train$Concrete_strength, new_test$`Superplasticizer `, new_test$Concrete_strength, iter = 1000, a = 0.1 )

lm5_new[1:3]
```
```{r}
#plot the data
plot(new_train$`Superplasticizer `, new_train$Concrete_strength, xlab = 'Superplasticizer',
       ylab ="strength")
lines(new_train$`Superplasticizer `, lm5_new$y_pred_train, col = "red")
```

##### 6. Coarse Aggregate
```{r}
lm6_new = lm(new_train$Coarse_Aggregate,new_train$Concrete_strength, new_test$Coarse_Aggregate, new_test$Concrete_strength, iter = 1000, a = 0.1 )

lm6_new[1:3]
```
```{r}
#plot the data
plot(new_train$Coarse_Aggregate, new_train$Concrete_strength, xlab = 'Coarse Aggregate',
       ylab ="Strength")
lines(new_train$Coarse_Aggregate, lm6_new$y_pred_train, col = "red")
```

##### 7. Fine Aggregate
```{r}
lm7_new = lm(new_train$Fine_Aggregate,new_train$Concrete_strength, new_test$Fine_Aggregate, new_test$Concrete_strength, iter = 1000, a = 0.1 )

lm7_new[1:3]
```

```{r}
#plot the data
plot(new_train$Fine_Aggregate, new_train$Concrete_strength, xlab = 'Fine Aggregate',
       ylab ="Strength")
lines(new_train$Fine_Aggregate, lm7_new$y_pred_train, col = "red")
```

##### 8. Age
```{r}
lm8_new = lm(new_train$Age,new_train$Concrete_strength, new_test$Age, new_test$Concrete_strength, iter = 1000, a = 0.1 )

lm8_new[1:3]
```
```{r}
#plot the data
plot(new_train$Age, new_train$Concrete_strength, xlab = 'Age',
       ylab ="Strength")
lines(new_train$Age, lm8_new$y_pred_train, col = "red")
```

```{r}
library(palmerpenguins)
par(mfrow=c(1,3))

#6
plot(new_train$Coarse_Aggregate, new_train$Concrete_strength, xlab = 'Coarse Aggregate',
       ylab ="Strength")
lines(new_train$Coarse_Aggregate, lm6_new$y_pred_train, col = "red")

#7
plot(new_train$Fine_Aggregate, new_train$Concrete_strength, xlab = 'Fine Aggregate',
       ylab ="Strength")
lines(new_train$Fine_Aggregate, lm7_new$y_pred_train, col = "red")

#8
plot(new_train$Age, new_train$Concrete_strength, xlab = 'Age',
       ylab ="Strength")
lines(new_train$Age, lm8_new$y_pred_train, col = "red")
```

### Multivariate linear regression

```{r}
#function for multivariate
mlm <- function(x_train, y_train, x_test, y_test, iter = 1000, alpha ){
  
  # Initialize parameters
  m <- rep(0, ncol(x_train))
  b <- 0
  
  n <- nrow(x_train)
  
  # calculate gradient descent
  for(i in 1:iter){
    y_pred <- x_train %*% m + b
    residuals <- y_train - y_pred
    grad_m <- (-2/n) * colSums(sweep(x_train, 2, residuals, FUN="*"))
    grad_b <- (-2/n) * sum(residuals) 
    m <- m - alpha * grad_m  # Update m
    b <- b - alpha * grad_b  # Update b
  }
  
  # Predictions
  y_pred_train <- x_train %*% m + b
  y_pred_test <- x_test %*% m + b
    
  # Calculate MSE, Variance and R-squared for training data
  MSE_train <- sum((y_train - y_pred_train)^2) / n
  Var_train <- sum((y_train - mean(y_train))^2) / n
  r_squared_train <- 1 - (MSE_train / Var_train)
  
  # Calculate MSE, Variance and R-squared for test data
  n_test = nrow(x_test)
  MSE_test <- sum((y_test - y_pred_test)^2) / n_test
  Var_test <- sum((y_test - mean(y_test))^2) / n_test
  r_squared_test <- 1 - (MSE_test / Var_test)

  # Return the coefficients, variance explained and predictions(for plot)
  list(coefficients = c(m = m, b = b), VE_train = r_squared_train, VE_test = r_squared_test)
}
```

```{r}
x_raw_train = as.matrix(raw_train[,1:8])
y_raw_train = as.matrix(raw_train[,9])
x_raw_test = as.matrix(raw_test[,1:8])
y_raw_test = as.matrix(raw_test[,9])

x_new_train = as.matrix(new_train[,1:8])
y_new_train = as.matrix(new_train[,9])
x_new_test = as.matrix(new_test[,1:8])
y_new_test = as.matrix(new_test[,9])
```


#### Raw Data
```{r, warning=FALSE}
# Convert the data input to matrix form
x_raw_train = as.matrix(raw_train[,1:8])
y_raw_train = as.matrix(raw_train[,9])
x_raw_test = as.matrix(raw_test[,1:8])
y_raw_test = as.matrix(raw_test[,9])

lm9_raw = mlm(x_raw_train, y_raw_train, x_raw_test, y_raw_test, iter = 1000, alpha = 0.000000001)

lm9_raw[1:3]
```

#### Pre-processed Data
```{r, warning=FALSE}
x_new_train = as.matrix(new_train[,1:8])
y_new_train = as.matrix(new_train[,9])
x_new_test = as.matrix(new_test[,1:8])
y_new_test = as.matrix(new_test[,9])

lm9_new = mlm(x_new_train, y_new_train, x_new_test, y_new_test, iter = 1000, alpha = 0.0004)

lm9_new[1:3]
```

#### Variance explained of models on training data 
```{r}
library(flextable)
library(magrittr)

#Variance explained of models on training data

raw_ve_train <-as.numeric(c(lm1_raw[2],lm3_raw[2],lm3_raw[2],lm4_raw[2],lm5_raw[2], lm6_raw[2],lm7_raw[2], lm8_raw[2],lm9_raw[2]))

new_ve_train <- as.numeric(c(lm1_new[2],lm2_new[2],lm3_new[2],lm4_new[2],lm5_new[2], lm6_new[2],lm7_new[2], lm8_new[2],lm9_new[2] ))

df_ve_train <- data.frame('Feature' = c("Cement",
                 "Blast Furnace Slag",
                 "Fly Ash",
                 "Water",
                 "Superplasticizer ",
                 "Coarse Aggregate",
                 "Fine Aggregate",
                 "Age",
                 "All 8 features"),
           'VE on train(raw)' = raw_ve_train,
           'VE on train(processed)' = new_ve_train
)



df_ve_train %>% regulartable() %>% autofit()

```

#### Variance explained of models on testing data 
```{r}
#Variance explained of models on testing data
raw_ve_test <- as.numeric(c(lm1_raw[3],lm3_raw[3],lm3_raw[3],lm4_raw[3],lm5_raw[3], lm6_raw[3],lm7_raw[3], lm8_raw[3],lm9_raw[3]))

new_ve_test <- as.numeric(c(lm1_new[3],lm2_new[3],lm3_new[3],lm4_new[3],lm5_new[3], lm6_new[3],lm7_new[3], lm8_new[3],lm9_new[3]))

df_ve_test <- data.frame('Feature' = c("Cement",
                 "Blast Furnace Slag",
                 "Fly Ash",
                 "Water",
                 "Superplasticizer ",
                 "Coarse Aggregate",
                 "Fine Aggregate",
                 "Age",
                 "All 8 features"),
                 'VE on test(raw)' = raw_ve_test,
                 'VE on test(processed)' = new_ve_test
)



df_ve_test %>% regulartable() %>% autofit()
```

### Compare and contrast models


```{r}
df1 <- data.frame('Feature' = c("Cement",
                 "Blast Furnace Slag",
                 "Fly Ash",
                 "Water",
                 "Superplasticizer ",
                 "Coarse Aggregate",
                 "Fine Aggregate",
                 "Age",
                 "All 8 features"),
           'VE on train(raw)' = raw_ve_train,
           'VE on test(raw)' = raw_ve_test,
           'VE on train(processed)' = new_ve_train,
           'VE on test(processed)' = new_ve_test
)



df1 %>% regulartable() %>% autofit()
```


